---
title: "Dataset Preparation"
author: Deepankar Vyas, 23200527
format:
  html:
    embed-resources: true
  pdf: default
execute: 
  warning: false
---

## DATASET PREPARATION

Here, we will web scrape data from `Football.co.uk[1]` website, for the last 9 seasons of the German league - Bundesliga. This dataset will help us provide various match stats for both Home and Away teams such as :-

1.  Goals
2.  Shots on Target
3.  Corners
4.  Fouls Commited
5.  Yellow and Red Cards
6.  Betting odds of various organizations

We will be using Goals, Shots on Target , Corners of both the Home and Away teams as well as Result of the matches. We will we scarp data of the past 9-10 seasons i.e (2012/13-2023/24)

```{r}
#| label: dataset preparation - loading libraries
#| echo: false

library(knitr)
library(rvest)
library(dplyr)
library(stringr)
library(lubridate)
library(tidyr)
```

```{r}
#| label : Dataset preparation-football.co.uk

# Base URL
base_url <- "https://www.football-data.co.uk/mmz4281"

# Seasons
seasons <- c("1213", "1314", "1415", "1516", "1617", "1718", "1819", "1920", "2021", "2122", "2223", "2324")

# Directory to save files
dir.create("Bundesliga_2012_2024", showWarnings = FALSE)
directory <- "Bundesliga_2012_2024"

# Function to download and read a CSV file
download_and_read_csv <- function(season) {
  file_url <- paste0(base_url, "/", season, "/D1.csv")
  destfile <- paste0(directory, "/D1_", season, ".csv")
  
  # Download the file
  download.file(file_url, destfile, mode = "wb")
  
  # Read the CSV file
  df <- read.csv(destfile)
  
  invisible (df)
}

# Download and combine all CSV files into a master dataset
master_dataset <- bind_rows(lapply(seasons, download_and_read_csv))

# Convert Date to Date type
master_dataset$Date <- dmy(master_dataset$Date)




```

Now, we will web scrap player and team's ratings from `FIFA Index[2]` website staring from FIFA 12 till FIFA 24. These are the ratings given by FIFA video game to teams and players and are generated periodically, taking into account current form of teams and players, and thus are a good measure of the current state of the team. We will first focus on the team ratings. We will web scrap following components of team ratings-

1.  Attack Rating
2.  Defense Rating
3.  Midfield Rating
4.  Overall Rating

```{r}
#| label: Dataset preparation-fifa_ratings_team

# Base URL components
base_url <- "https://fifaindex.com/teams/fifa24"
league_params <- "league=19&league=20&order=desc"
max_pages <- 3

# Function to scrape data from a given URL
scrape_fifa_data <- function(url) {
  page <- try(read_html(url), silent = TRUE)
  if (inherits(page, "try-error")) {
    return(data.frame()) # Return empty data frame if error occurs
  }
  
  # Extract league names
  leagues <- page %>% html_nodes(".link-league") %>% html_text(trim = TRUE)
  
  # Validate league of the first two teams
  if (length(leagues) < 2 || !all(leagues[1:2] %in% c("Bundesliga", "Germany 1. Bundesliga (1)", "Bundesliga 2", "Germany 2. Bundesliga (2)"))) {
    return(data.frame()) # Return empty data frame if the first two teams' leagues are not Bundesliga
  }
  
  # Extract team names
  teams <- page %>% html_nodes(".link-team") %>% html_text(trim = TRUE)
  teams <- teams[seq(2,length(teams),2)]
  
  # Extract ratings
  att <- page %>% html_nodes("td[data-title='ATT'] span") %>% html_text(trim = TRUE)
  mid <- page %>% html_nodes("td[data-title='MID'] span") %>% html_text(trim = TRUE)
  def <- page %>% html_nodes("td[data-title='DEF'] span") %>% html_text(trim = TRUE)
  ovr <- page %>% html_nodes("td[data-title='OVR'] span") %>% html_text(trim = TRUE)
  
  if (length(att) == 0 || length(mid) == 0 || length(def) == 0 || length(ovr) == 0) {
    return(data.frame()) # Return empty data frame if no ratings found
  }
  
  date <- page %>% html_nodes("a.dropdown-toggle") %>% html_text(trim = TRUE) %>% .[6]

  
  data <- data.frame(
    Team = teams,
    Date = date,
    Attack = as.numeric(att),
    Midfield = as.numeric(mid),
    Defense = as.numeric(def),
    Overall = as.numeric(ovr),
    stringsAsFactors = FALSE
  )
  
  return(data)
}

# Initialize an empty data frame to store the results
all_fifa_data <- data.frame()

# Loop through all editions and their possible date values
  for (id in 8:612) {
    for (page in 1:max_pages) {
      # Construct URL
      if (page == 1) {
        url <- paste0(base_url, "_", id, "/?", league_params)
      } else {
        url <- paste0(base_url, "_", id, "/?page=", page, "&", league_params)
      }
      
      # Scrape data from the URL
      data <- scrape_fifa_data(url)
      if (nrow(data) == 0) break # If no data, exit the page loop
      
      # Append to the master dataset
      all_fifa_data <- bind_rows(all_fifa_data, data)
    }

  }

#Scraping for June 4, 2024
url1 = "https://fifaindex.com/teams/?league=19&league=20&order=desc"
url2 = "https://fifaindex.com/teams/?page=2&league=19&league=20&order=desc"

for (url in c(url1, url2)){
  
      data <- scrape_fifa_data(url)
      if (nrow(data) == 0) {
        }# If no data, do nothing
      
      else{
        
        all_fifa_data <- bind_rows(all_fifa_data, data)
      }
}

# Function to standardize month abbreviations
standardize_month_abbreviations <- function(date_string) {
  date_string <- str_replace_all(date_string, "Sept\\.", "Sep.")
  return(date_string)
}

all_fifa_data$Date <- sapply(all_fifa_data$Date, standardize_month_abbreviations)

# Convert Date to Date type
all_fifa_data$Date=mdy(all_fifa_data$Date)

# Ordering by date
all_fifa_data <- all_fifa_data |> arrange(desc(Date))



```

```{r}
#| label : Heuristic to combine team names

unique_teams_master <- unique(c(master_dataset$HomeTeam, master_dataset$AwayTeam))

# Define a mapping of alternative team names to standardized team names in master_dataset
team_name_mapping <- list(
  "Bayern Munich" = c("Bayern München", "FC Bayern München", "FC Bayern Munich"),
  "Dortmund" = c("Borussia Dortmund"),
  "RB Leipzig" = c("RB Leipzig"),
  "Leverkusen" = c("Bayer 04 Leverkusen", "Leverkusen"),
  "Ein Frankfurt" = c("Eintracht Frankfurt", "Frankfurt"),
  "Hoffenheim" = c("TSG 1899 Hoffenheim", "TSG Hoffenheim", "1899 Hoffenheim"),
  "Wolfsburg" = c("VfL Wolfsburg"),
  "Freiburg" = c("SC Freiburg", "Sport-Club Freiburg"),
  "Union Berlin" = c("1. FC Union Berlin", "Union Berlin"),
  "M'gladbach" = c("Borussia Mönchengladbach", "Borussia M'gladbach", "M'gladbach", "Borussia Mönchengladbach"),
  "Mainz" = c("1. FSV Mainz 05"),
  "Werder Bremen" = c("Werder Bremen", "SV Werder Bremen"),
  "Augsburg" = c("FC Augsburg"),
  "Stuttgart" = c("VfB Stuttgart"),
  "FC Koln" = c("1. FC Köln"),
  "Bochum" = c("VfL Bochum", "VfL Bochum 1848"),
  "Hertha" = c("Hertha BSC", "Hertha Berlin", "Hertha BSC Berlin"),
  "Heidenheim" = c("1. FC Heidenheim", "1. FC Heidenheim 1846", "Heidenheim"),
  "Hamburg" = c("Hamburger SV"),
  "Schalke 04" = c("FC Schalke 04"),
  "Darmstadt" = c("SV Darmstadt 98"),
  "Hannover" = c("Hannover 96"),
  "Fortuna Dusseldorf" = c("Fortuna Düsseldorf", "Düsseldorf"),
  "1. FC Kaiserslautern" = c("1. FC Kaiserslautern", "Kaiserslautern"),
  "Paderborn" = c("SC Paderborn 07"),
  "Karlsruher SC" = c("Karlsruher SC"),
  "FC St. Pauli" = c("FC St. Pauli"),
  "Nurnberg" = c("1. FC Nürnberg", "1. FC Nuremberg"),
  "Hansa Rostock" = c("Hansa Rostock"),
  "1. FC Magdeburg" = c("1. FC Magdeburg"),
  "Holstein Kiel" = c("Holstein Kiel"),
  "Greuther Furth" = c("SpVgg Greuther Fürth", "Fürth"),
  "Braunschweig" = c("Eintracht Braunschweig", "Braunschweig"),
  "VfL Osnabrück" = c("VfL Osnabrück"),
  "SV Elversberg" = c("SV Elversberg"),
  "SV Wehen Wiesbaden" = c("SV Wehen Wiesbaden"),
  "Bielefeld" = c("Arminia Bielefeld", "DSC Arminia Bielefeld"),
  "SV Sandhausen" = c("SV Sandhausen", "SV Sandhausen 1916"),
  "Jahn Regensburg" = c("Jahn Regensburg", "SSV Jahn Regensburg"),
  "Würzburger Kickers" = c("Würzburger Kickers", "FC Würzburger Kickers"),
  "Dynamo Dresden" = c("Dynamo Dresden", "SG Dynamo Dresden"),
  "FSV Frankfurt" = c("FSV Frankfurt", "FSV Frankfurt 1899"),
  "1860 München" = c("1860 München", "TSV 1860 München", "1860 Munich"),
  "FC Energie Cottbus" = c("FC Energie Cottbus", "Energie Cottbus"),
  "Alemannia Aachen" = c("Alemannia Aachen"),
  "Ingolstadt" = c("FC Ingolstadt 04")
)

# Reverse the mapping for quick lookup
reverse_mapping <- setNames(rep(names(team_name_mapping), lengths(team_name_mapping)), unlist(team_name_mapping))

# Standardize team names in all_fifa_data
all_fifa_data$Team <- unname(reverse_mapping[all_fifa_data$Team])

# Filter records in all_fifa_data to keep only those teams that are present in master_dataset
all_fifa_data <- all_fifa_data %>% filter(Team %in% unique_teams_master)
```

Now, we will Aggregate Ratings by Month and Year, so that each team has an Attack, Midfield, Defense and Overall Rating for each month of the years (2012-2024)

```{r}
#| label: Dataset-preparation-Aggregate_Ratings_by_Month_and_Year


# Extract Year and Month from Date
all_fifa_data <- all_fifa_data %>%
  mutate(Year = year(Date), Month = month(Date))

# Calculate average ratings by Team, Year, and Month
avg_ratings <- all_fifa_data %>%
  group_by(Team, Year, Month) %>%
  summarise(
    Avg_Attack = round(mean(Attack, na.rm = TRUE)),
    Avg_Midfield = round(mean(Midfield, na.rm = TRUE)),
    Avg_Defense = round(mean(Defense, na.rm = TRUE)),
    Avg_Overall = round(mean(Overall, na.rm = TRUE))
  ) %>%
  ungroup()



```

```{r}

# Ensure avg_ratings has continuous month-year data for each team
all_months <- expand.grid(Year = unique(avg_ratings$Year), Month = 1:12)
all_teams_months <- expand.grid(Team = unique(avg_ratings$Team), Year = unique(avg_ratings$Year), Month = 1:12)

avg_ratings_full <- all_teams_months %>%
  left_join(avg_ratings, by = c("Team", "Year", "Month")) %>%
  arrange(Team, Year, Month)

# Fill in missing values with the nearest previous available values
avg_ratings_full <- avg_ratings_full %>%
  group_by(Team) %>%
  fill(Avg_Attack, Avg_Midfield, Avg_Defense, Avg_Overall, .direction = "down") %>%
  ungroup()

# Removing NA values 
avg_ratings_full <- avg_ratings_full %>%
  filter(!(Year == 2011 & Month < 9))

avg_ratings_full <- avg_ratings_full %>%
  group_by(Team) %>%
  fill(Avg_Attack, Avg_Midfield, Avg_Defense, Avg_Overall, .direction = "up") %>%
  ungroup()



# Extract Year and Month from Date in master_dataset
master_dataset <- master_dataset %>%
  mutate(Year = year(Date), Month = month(Date))

# Join avg_ratings_full with master_dataset for HomeTeam
master_dataset <- master_dataset %>%
  left_join(avg_ratings_full, by = c("HomeTeam" = "Team", "Year", "Month")) %>%
  rename(
    Home_Attack = Avg_Attack,
    Home_Midfield = Avg_Midfield,
    Home_Defense = Avg_Defense,
    Home_Overall = Avg_Overall
  )

# Join avg_ratings_full with master_dataset for AwayTeam
master_dataset <- master_dataset %>%
  left_join(avg_ratings_full, by = c("AwayTeam" = "Team", "Year", "Month"))  %>% 
  rename(
    Away_Attack = Avg_Attack,
    Away_Midfield = Avg_Midfield,
    Away_Defense = Avg_Defense,
    Away_Overall = Avg_Overall
  )

# Remove extra columns
master_dataset <- master_dataset %>%
  select(-Year, -Month)

# Save the final master dataset to a CSV file
write.csv(master_dataset, "Bundesliga_Master_Dataset_Final.csv", row.names = FALSE)

```

## References

```{r}

```
